\documentclass[a4paper,UKenglish]{lipics}

\usepackage{microtype}
%\usepackage[colorinlistoftodos]{todonotes} 
\usepackage{tikz}
\usepackage{latexsym}     
\usepackage{hyperref}
\usepackage{multicol}
\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{textcomp}
\usepackage{bussproofs}

\usetikzlibrary{shapes}                
\usetikzlibrary{fit}

\bibliographystyle{plain}
\pagestyle{headings}

\title{Verification of redecoration for infinite triangular matrices
  using coinduction}

\author[1]{Ralph Matthes}
\author[2]{Celia Picard}
\affil[1]{Institut de Recherche en Informatique de Toulouse (IRIT), \\
  C.N.R.S. and University of Toulouse, France}
\affil[2]{Institut de Recherche en Informatique de Toulouse (IRIT), \\
  University of Toulouse, France}
\authorrunning{ R. Matthes and C. Picard} 

\subjclass{F.3.1 Specifying and Verifying and Reasoning about Programs}
\keywords{nested datatype, coinduction, theorem proving, Coq}\Copyright[nc-nd]{R. Matthes and C. Picard}
%%% before exchange with the editor, we had
%%\Copyright[nc-nd] % or nd only!
%%\phantom{}
\serieslogo{}%
\volumeinfo%
     {Nils Anders Danielsson, Bengt Nordstr\"om}%
     {2}%
     {18th International Workshop on Types for Proofs and Programs (TYPES~2011)}%
     {19}%
     {1}%
     {57}%
   \EventShortName{TYPES 2011}
   \DOI{10.4230/LIPIcs.TYPES.2011.57}%


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Macros for references
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\rdef}[1]{Definition~\ref{def:#1}}
\newcommand{\rlem}[1]{Lemma~\ref{lemma:#1}}
\newcommand{\rthm}[1]{Theorem~\ref{thm:#1}}
\newcommand{\rprop}[1]{Property~\ref{prop:#1}}
\newcommand{\rsect}[1]{Section~\ref{sect:#1}}
\newcommand{\rfig}[1]{Figure~\ref{fig:#1}}
\newcommand{\rcor}[1]{Corollary~\ref{cor:#1}}
\newcommand{\rex}[1]{Example~\ref{ex:#1}}
\newcommand{\rrem}[1]{Remark~\ref{rem:#1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Macros for Tri
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Tri}{\ensuremath{\mathit{Tri}}}
\newcommand{\Trap}{\ensuremath{\mathit{Trap}}}
\newcommand{\pair}[2]{\langle #1,#2\rangle}
\newcommand{\redec}{\ensuremath{\mathit{redec}}}
\newcommand{\Rel}{\ensuremath{\mathit{Rel}}}
\newcommand{\Prop}{\ensuremath{\mathit{Prop}}}
\newcommand{\bisim}{\ensuremath{\mathbin{\simeq}}} %~~
\newcommand{\topT}{\ensuremath{\mathit{top}}}
\newcommand{\rest}{\ensuremath{\mathit{rest}}}
\newcommand{\cut}{\ensuremath{\mathit{cut}}}
\newcommand{\escut}{\ensuremath{\mathit{es\_cut}}}
\newcommand{\frow}{\ensuremath{\mathit{frow}}}
\newcommand{\constr}{\ensuremath{\mathit{constr}}}
\newcommand{\lift}{\ensuremath{\mathit{lift}}}
\newcommand{\fst}{\ensuremath{\mathit{fst}}}

\newcommand{\counit}{\ensuremath{\mathit{counit}}}
\newcommand{\cobind}{\ensuremath{\mathit{cobind}}}
\newcommand{\simcom}{\ensuremath{\mathbin{\approxeq}}} %~~^
\newcommand{\addes}{\ensuremath{\mathit{addes}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Macros for Stream
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Str}{\ensuremath{\mathit{Str}}}
\newcommand{\EqSt}{\ensuremath{\mathbin{\equiv}}} %==
\newcommand{\map}{\ensuremath{\mathit{map}}}
\newcommand{\tl}{\ensuremath{\mathit{tl}}}
\newcommand{\hd}{\ensuremath{\mathit{hd}}}
\newcommand{\zipWith}{\ensuremath{\mathit{zipWith}}}
\newcommand{\es}{\ensuremath{\mathit{es}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Macros for TriS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\vari}{'}%{^\star}
\newcommand{\TriS}{\ensuremath{\mathit{Tri\vari}}}
\newcommand{\tSR}{\ensuremath{\mathit{toStreamRep}}}
\newcommand{\fSR}{\ensuremath{\mathit{fromStreamRep}}}
\newcommand{\bisimS}{\ensuremath{\mathbin{\cong}}} %==^
\newcommand{\topS}{\ensuremath{\mathit{top\vari}}}
\newcommand{\frowS}{\ensuremath{\mathit{frow\vari}}}
\newcommand{\restS}{\ensuremath{\mathit{rest\vari}}}
\newcommand{\redecS}{\ensuremath{\mathit{redec\vari}}}
\newcommand{\liftS}{\ensuremath{\mathit{lift\vari}}}
\newcommand{\redecSG}{\ensuremath{\mathit{redec\vari_{gen}}}}
\newcommand{\redecSS}{\ensuremath{\mathit{redec\vari_{alt}}}}
\newcommand{\addesS}{\ensuremath{\mathit{addes\vari}}}
\newcommand{\tls}{\ensuremath{\mathit{tails}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Macros for Tri_fin
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Trif}{\ensuremath{\mathit{Tri_{fin}}}}
\newcommand{\constrf}{\ensuremath{\mathit{constr_{fin}}}}
\newcommand{\sgf}{\ensuremath{\mathit{sg_{fin}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Macros for TriL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\cutL}{\ensuremath{\mathit{cutL}}}
\newcommand{\tLR}{\ensuremath{\mathit{toListRep}}}
\newcommand{\fLR}{\ensuremath{\mathit{fromListRep}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

\maketitle



\begin{abstract}
Finite triangular matrices with a dedicated type for the diagonal
elements can be profitably represented by a nested data type, i.\,e., a
heterogeneous family of inductive data types, while infinite
triangular matrices form an example of a nested coinductive type,
which is a heterogeneous family of coinductive data types. 

Redecoration for infinite triangular matrices is taken up from
previous work involving the first author, and it is shown that
redecoration forms a comonad with respect to bisimilarity.

The main result, however, is a validation of the original algorithm
against a model based on infinite streams of infinite streams. The
two formulations are even provably equivalent, and the second is
identified as a special instance of the generic cobind operation
resulting from the well-known comultiplication operation on streams
that creates the stream of successive tails of a given stream. Thus,
perhaps surprisingly, the verification of redecoration is easier for
infinite triangular matrices than for their finite counterpart.

All the results have been obtained and are fully formalized in the
current version of the Coq 
theorem proving environment where these coinductive datatypes are
fully supported since the version 8.1, released in 2007. 
Nonetheless, instead of displaying the Coq development, we have chosen
to write the paper in standard mathematical and type-theoretic 
language. Thus, it should be accessible without any specific knowledge about
Coq.

\end{abstract}
\input{intro}
\input{original}
\input{streams}
\input{conclusion}

%\bibliography{coredec}

\begin{thebibliography}{10}

\bibitem{grossestcspaper}
Andreas Abel, Ralph Matthes, and Tarmo Uustalu.
\newblock Iteration and coiteration schemes for higher-order and nested
  datatypes.
\newblock {\em Theoretical Computer Science}, 333(1--2):3--66, 2005.

\bibitem{haskellList:redecorationAnswer}
Jorge Adriano.
\newblock Answer to {M}arkus {S}chnell's message \texttt{slide: useful
  function?}
\newblock {Haskell Mailing List}, November 2002.

\bibitem{coqart}
Yves Bertot and Pierre Cast{\'e}ran.
\newblock {\em Interactive Theorem Proving and Program Development.
  {C}oq'{A}rt: The Calculus of Inductive Constructions}.
\newblock Texts in Theoretical Computer Science. Springer Verlag, 2004.

\bibitem{birdmeertens}
Richard Bird and Lambert Meertens.
\newblock Nested datatypes.
\newblock In Johan Jeuring, editor, {\em Mathematics of {P}rogram
  {C}onstruction, {MPC}'98, Proceedings}, volume 1422 of {\em Lecture Notes in
  Computer Science}, pages 52--67. Springer Verlag, 1998.

\bibitem{gfolds}
Richard Bird and Ross Paterson.
\newblock Generalised folds for nested datatypes.
\newblock {\em {F}ormal {A}spects of {C}omputing}, 11(2):200--222, 1999.

\bibitem{dan}
Nils~Anders Danielsson.
\newblock Beating the productivity checker using embedded languages.
\newblock In Ana Bove, Ekaterina Komendantskaya, and Milad Niqui, editors, {\em
  PAR}, volume~43 of {\em EPTCS}, pages 29--48, 2010.

\bibitem{daal}
Nils~Anders Danielsson and Thorsten Altenkirch.
\newblock Subtyping, declaratively.
\newblock In Claude Bolduc, Jules Desharnais, and B{\'e}chir Ktari, editors,
  {\em Mathematics of Program Construction}, volume 6120 of {\em Lecture Notes
  in Computer Science}, pages 100--118. Springer, 2010.

\bibitem{maclane}
Saunders {Mac Lane}.
\newblock {\em Categories for the Working Mathematician}, volume~5 of {\em
  Graduate Texts in Mathematics}.
\newblock Springer Verlag, second edition, 1998.

\bibitem{TYPES11code}
Ralph Matthes and Celia Picard.
\newblock Formalization in {C}oq for this article, 2012.
\newblock \url{www.irit.fr/~Celia.Picard/Coq/Redecoration/}.

\bibitem{types07}
Ralph Matthes and Martin Strecker.
\newblock Verification of the redecoration algorithm for triangular matrices.
\newblock In Furio Honsell, Marino Miculan, and Ivan Scagnetto, editors, {\em
  Types for Proofs and Programs, International Conference, TYPES 2007, Revised
  Selected Papers}, volume 4941 of {\em Lecture Notes in Computer Science},
  pages 125--141. Springer Verlag, 2008.

\bibitem{theseCP}
Celia Picard.
\newblock {\em Représentation coinductive des graphes}.
\newblock PhD thesis, Université de Toulouse, 2012.

\bibitem{DBLP:conf/tphol/SozeauO08}
Matthieu Sozeau and Nicolas Oury.
\newblock First-class type classes.
\newblock In Otmane~A\"{\i}t Mohamed, C{\'e}sar Mu{\~n}oz, and Sofi{\`e}ne
  Tahar, editors, {\em TPHOLs}, volume 5170 of {\em Lecture Notes in Computer
  Science}, pages 278--293. Springer, 2008.

\bibitem{MC}
{T}he {C}oq~{D}evelopment {T}eam.
\newblock The {C}oq {P}roof {A}ssistant {R}eference {M}anual.
\newblock INRIA.

\bibitem{DBLP:conf/sfp/UustaluV01}
Tarmo Uustalu and Varmo Vene.
\newblock The dual of substitution is redecoration.
\newblock In Kevin Hammond and Sharon Curtis, editors, {\em Scottish Functional
  Programming Workshop}, volume~3 of {\em Trends in Functional Programming},
  pages 99--110. Intellect, 2001.

\end{thebibliography}


\end{document}
